Improving Adaptive Seamless Designs through Bayesian optimization

### 2021-08-28 response_to_the_reviewers2.md

---
title: Response to the Reviewers
author: Jakob Richter
header-includes: |
  \usepackage{framed}
  \newlength{\leftbarwidth}
  \setlength{\leftbarwidth}{3pt}
  \newlength{\leftbarsep}
  \setlength{\leftbarsep}{-10pt}
  \newcommand*{\leftbarcolorcmd}{\color{leftbarcolor}}
  \colorlet{leftbarcolor}{gray}
  \renewenvironment{leftbar}{%
      \def\FrameCommand{{\leftbarcolorcmd{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
      \MakeFramed {\advance \hsize -\width \FrameRestore }%
    \noindent
  }{%
      \endMakeFramed
  }
  \renewenvironment{quote}
  {\vspace{5pt} \leftbar\list{}{\rightmargin\leftmargin}%
  \item\vspace{-5pt}\relax}
    {\endlist\endleftbar}
---

# Second response to the reviews for "Improving Adaptive Seamless Designs through Bayesian optimization"

We would like to thank all reviewers again for their valuable and constructive comments on the paper.
We carried out a new simulation to address the comments by Reviewer 2 and included the results in the appendix.

## Reviewer: 1

### Comments to the Author

Thanks for addressing/responding to all my comments

## Reviewer: 2

### Comments to the Author
Thanks authors for the response to the comments. However, some of them are not addressed appropriately here. Please consider them carefully in the next round revision.
For page 7 table 1 effect size scenario, although it is common that early come is overoptimistic and higher, however, it is also often in practice that there are delayed treatment effect (e.g. in the immunotherapy clinical trial setting) where stage 1 effect are smaller than stage 2. I would strongly recommend authors to add difference simulation setting.

> We carried out new simulations to address this comment. 
> In the desired scenario, titled `paper_rev`, effect sizes for early and final are flipped compared to the standard scenario `paper`, so that the lower values occur in the early stage.
> The results are shown in the Appendix.

Similarly for the selection of “espilon” and “threshold” in page 8 table 2, there should be more practical evaluation where they are commonly based on the clinical meaningful cut, besides as optimization parameters. Suggest to add relevant discussion here to facilitate more practical usage of proposed methods.

> In this paper we consider an optimization problem maximizing the power given a total sample size.
> If, for example, the task was to select all treatments with clinically relevant effect sizes, then clearly the selection parameter needs to be informed by clinical reasoning.
> We added a respective statement to the discussion.

Page 8 table 3, authors should add more details or rationale on the how parameters range of search space is determined.

> We included a statement that “epsilon” and “threshold” are obtained through preliminary studies.
> Furthermore, we added a Figure S4 in the Appendix, which demonstrates that the ranges are chosen in a sensible way.

### Minor comments:
There is citation format typo in page 3: “as implemented in the asd package (?)”

> This occurred only in the diff-version, which will not be published.

### 2021-08-28 main.tex
jr+tf

The goal of this simulation study is to investigate the potential benefit of BO for efficiently finding a clinical trial design that maximizes a given metric.
The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.
Depending on the application the power can be calculated as the proportion of correctly rejecting any, all, or a subset of the elementary hypotheses~\citep{senn_power_2007}.

\subsection{Clinical trial design and parameters}
\label{subsec:design}
For the true effect sizes in the simulation, we consider four \emph{effect sets} whose values are inspired and partially based on the corresponding numbers of the motivating example, see Table~\ref{tab:table_effect_names}.
\input{generated/tables/table_effect_names.tex}
For the effect set \emph{paper} we use exactly the same numbers as in~\citet{friede_adaptive_2020}.
For the effect set \emph{paper2} the effect sizes for the second stage are doubled.
As further case we consider a \emph{linear} effect set with effect sizes linear increasing from 20\% to 80\% for the first stage and effect sizes divided by 4 for the second stage as well as a \emph{sigmoid} relationship, again with 4 times larger effect sizes for the first stage.
These sets reflect different realistic situations.
A further effect set, where the numbers between the early and the final stage are flipped for \emph{paper} is presented in the Appendix.

The strategies for treatment selection for the second stage in the two-phase trial design and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
Choices for the boundaries of $\epsilon$ and $\tau$ are obtained from preliminary studies and based on the fact that these are thresholds to the absolute amount of normally distributed random variables.
Also Figure~\ref{fig:plot_appendix_opt_episolon_thresh} in the Appendix shows that the optimum is within the given boundaries.
The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
The $\kappa$-best strategies select the 1, 2, 3 or 4 best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). 
The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of the total samples (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation       & Procedure of selection strategy  \\
  \hline
  \emph{$\kappa$-best}    & Select $\kappa = 1, 2, 3$ or 4 (all) arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
  \emph{thresh}  & Select all arms with test statistic larger than the value $\tau$ (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter                        & Range \\
  \hline
  Selection strategy               & \emph{\{1-best, 2-best, 3-best, all, eps, thresh\}} \\
  $r$ (ratio)                      & $(0,1)$ \\
  $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
  $\tau$ (only for \emph{thresh})  & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

In this paper we analyzed if BO can quickly find an adaptive seamless design with high power, more efficiently than an exhaustive grid search.
For various sets of treatment effect sizes and total sample sizes, BO was able to find competitive designs in a fraction of the time.
In contrast to grid search, BO was able to account for stochasticity in the metric that should be optimized.
%In addition to results in other research fields, our results underline that BO is a powerful method for finding optimal configurations for problems that fit the black-box definition.
For our optimization we chose wide intervals for the parameters of the selection rules (i.e. epsilon and threshold rule) to show that BO is generally applicable.
In practical applications, however, these values might sometimes be based on clinical rather than statistical considerations or the range of these values might be restricted by such considerations.
In general, the application of BO to clinical trial design selection in general seems promising.
Additional restrictions on the search space decrease the its size and therefore promise even more reliable results.
The same concept could be applied for subgroup selection in adaptive enrichment designs, which make use of interim data in selecting the target population for the remainder of the trial~\citep{burnett_adaptive_2020}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_appendix_boxplot_paper_rev.pdf}
\caption{%
    Box plots that compares the performance (power) values on the effect set \emph{paper} and \emph{paper\_rev}.
    \emph{paper\_rev} is an additional scenario, where the effects for the stage \emph{early} and \emph{final} are flipped.
  In contrast to \emph{paper}, for \emph{paper\_rev} the effect sizes are smaller for the \emph{early} stage.
}
\label{fig:plot_appendix_boxplot_paper_rev} 
\end{figure}

>>>>

The goal of this simulation study is to investigate the potential benefit of BO for efficiently finding a clinical trial design that maximizes a given metric.
The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
In the first stage, five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.
Depending on the application the power can be calculated as the proportion of correctly rejecting any, all, or a subset of the elementary hypotheses~\citep{senn_power_2007}.

\subsection{Clinical trial design and parameters}
\label{subsec:design}
For the true effect sizes in the simulation, we consider four \emph{effect sets} whose values are inspired and partially based on the corresponding numbers of the motivating example, see Table~\ref{tab:table_effect_names}.
\input{generated/tables/table_effect_names.tex}
For the effect set \emph{paper} we use exactly the same numbers as in~\citet{friede_adaptive_2020}.
For the effect set \emph{paper2} the effect sizes for the second stage are doubled.
As a further case, we consider a \emph{linear} effect set with effect sizes linear increasing from 20\% to 80\% for the first stage and effect sizes divided by 4 for the second stage as well as a \emph{sigmoid} relationship, again with 4 times larger effect sizes for the first stage.
These sets reflect different realistic situations.
Sometimes, there are delayed treatment effects (e.g.\ in the immunotherapy clinical trial setting) where the early effects are smaller than the delayed final effects.
To include this scenario, we added a further effect set, where the numbers between the early and the final stage of the scenario \emph{paper} are flipped (see Appendix).

The strategies for treatment selection for the second stage in the two-phase trial design and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
Choices for the boundaries of $\epsilon$ and $\tau$ are obtained from preliminary studies and based on the fact that these are thresholds to the absolute amount of normally distributed random variables.
Also, Figure~\ref{fig:plot_appendix_opt_episolon_thresh} in the Appendix shows that the optimum is within the given boundaries.
The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
The $\kappa$-best strategies select the 1, 2, 3 or 4 best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). 
The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of the total samples (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation       & Procedure of selection strategy  \\
  \hline
  \emph{$\kappa$-best}    & Select $\kappa = 1, 2, 3$ or 4 (all) arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
  \emph{thresh}  & Select all arms with test statistic larger than the value $\tau$ (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter                        & Range \\
  \hline
  Selection strategy               & \emph{\{1-best, 2-best, 3-best, all, eps, thresh\}} \\
  $r$ (ratio)                      & $(0,1)$ \\
  $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
  $\tau$ (only for \emph{thresh})  & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

In this paper we analyzed if BO can quickly find an adaptive seamless design with high power, more efficiently than an exhaustive grid search.
For various sets of treatment effect sizes and total sample sizes, BO was able to find competitive designs in a fraction of the time.
In contrast to grid search, BO was able to account for stochasticity in the metric that should be optimized.
%In addition to results in other research fields, our results underline that BO is a powerful method for finding optimal configurations for problems that fit the black-box definition.
For our optimization we chose wide intervals for the parameters of the selection rules (i.e. epsilon and threshold rule) to show that BO is generally applicable.
In practical applications, however, these values might sometimes be based on clinical rather than statistical considerations or the range of these values might be restricted by such considerations.
Here we consider an optimization problem maximizing the power given a total sample size.
If, for example, the task was to select all treatments with clinically relevant effect sizes, then clearly the selection parameter needs to be informed by clinical reasoning.
In general, the application of BO to clinical trial design selection in general seems promising.
Additional restrictions on the search space decrease its size and therefore promise even more reliable results.
The same concept could be applied for subgroup selection in adaptive enrichment designs, which make use of interim data in selecting the target population for the remainder of the trial~\citep{burnett_adaptive_2020}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_appendix_boxplot_paper_rev.pdf}
\caption{%
    Box plots that compare the performance (power) values on the effect set \emph{paper} and \emph{paper\_rev}.
    \emph{paper\_rev} is an additional scenario, where the effects for the stage \emph{early} and \emph{final} are flipped.
  In contrast to \emph{paper}, for \emph{paper\_rev} the effect sizes are smaller for the \emph{early} stage.
}
\label{fig:plot_appendix_boxplot_paper_rev} 
\end{figure}
